---
title: "Lab 02"
subtitle: "CS3172-1, Spring 2023, Effat University"
author: "farha hussaini - S20106851"
output: pdf_document
---

# Packages

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse) 
library(scales)
library(ggplot2)
``` 

# Data

```{r read-in-data, message = FALSE}
cas <- read_rds("data/canada_survey.rds")
```

```{r}
dim(cas)
```



# Tasks

## Task 1

1.	Task 1: Data cleaning:
•	Subset `cas` so that variables `energy_expense` and  `household_income` only contain values greater than 0. Overwrite  `cas`.
•	Use function `factor()` to change the variable `marital_status` to  be a factor rather than double. Overwrite `cas`. Consult the data  dictionary and write-out what the marital status codes mean.


```{r}
cas <-filter( cas, energy_expense > 0, household_income > 0)

```
```{r}
cas <- mutate(cas,marital_status = factor(marital_status))
```


## Task 2

2.	Task 2: Variable Recoding
Use function `case_when()` from `dplyr` to recode the two variables below. Overwrite `cas` after each recoding. Do not use function `if_else()` to complete this task.
Here is a small example of `case_when()` in action. Consult R’s help for further quizples.

Recode `heat_equip` so instead of having values 1, 2, 3, 4, it contains values `"steam"`, `"forced air"`, `"stove"`, and `"electric heating"` according to the data dictionary. These new values are as defined below:
o	`steam`: steam or water furnace
o	`forced air`: forced air furnace
o	`stove`: heating stoves, cookstove, or other
o	`electric heating`: electric

Recode `heat_fuel` so instead of having values 1, 2, 3, 4, it contains values `"oil"`, `"gas"`, `"electricity"`, and `"other"` according to the data dictionary. These new values are as defined below:
o	`oil`: oil or other liquid fuel
o	`gas`: natural gas
o	`electricity`: electricity
o	`other`: bottled gas, wood, or other


```{r}
cas <- mutate(cas, heat_euip = case_when(heat_equip == 1 ~ 'steam',
                                         heat_equip == 2 ~ 'forced air ',
                                         heat_equip == 3 ~ 'stove',
                                         heat_equip == 4 ~ 'electric heating'))
```

```{r}
cas <- mutate(cas, heat_euip = case_when(heat_fuel == 1 ~ "oil",
  heat_fuel == 2 ~ "gas",
  heat_fuel == 3 ~ "electricity",
  heat_fuel == 4 ~ "other"))
```

## Task 3

3.	Task 3: Group_by and Summarize
For each combination of heating fuel type and heating equipment, find the mean, median, and standard deviation of household energy expenditures. Print your results.

Based on your findings, answer the following questions in 1 - 2 sentences total.
o	What combination of fuel type and equipment has the highest average energy expense?
o	Which combination has the most variability with regards to energy expense?
o	Which type of heating equipment doesn’t take all possible fuel types?



```{r}
cas1 <- cas %>% 
  group_by(heat_fuel, heat_equip) %>% 
 summarize(
    mean_energy_expense = mean(energy_expense),
    median_energy_expense = median(energy_expense),
    sd_energy_expense = sd(energy_expense)
  ) %>% 
  print()
```


- Based on the results, we can answer the following questions:
- The combination of heating fuel type "electricity" and heating equipment "electric heating" has the highest average energy expense.
- The combination of heating fuel type "electricity" and heating equipment "stove" has the most variability with regards to energy expense, as it has the highest standard deviation.
- Heating equipment "steam" doesn't take all possible fuel types, as it is only used with heating fuel type "oil".

## Task 4

4.	Task 4: Data Visualization
Create a bar chart of energy expenses by heating fuel type and faceted by the type of heating equipment in a 2 x 2 grid. Your axis should be appropriately labeled with a dollar sign and commas. The `scales` package may be helpful here.



```{r}
ggplot(cas, mapping = aes (x= heat_fuel, y = energy_expense)) + 
  geom_boxplot() + 
  scale_y_continuous(labels = scales ::dollar_format()) + 
  facet_wrap(~ heat_euip, nrow = 2) +
  labs(title = "energy expense vs heating type",
       x = "heat fuel type",
       y = "energy expense")

```


## Task 5

5.	Task 5: Mutate()
Create a new variable describing the proportion of household income spent on energy related expenses, and then find the respondent that spent the highest proportion of their household income on energy and the respondent that spent the lowest proportion of their household income on energy. End your pipeline with the tibble being passed into `glimpse()`. Describe these respondents based on the data they have provided.


```{r}
cas2 <- cas %>% mutate (energy_prop = energy_expense/household_income) %>%
  arrange(desc(energy_prop)) %>%
  slice(1,n()) %>%
  glimpse()
```

## Task 6

6.	Task 6: Pipeline
For each year, find the province with the cheapest median energy expense per room. Your answer should consist of a single `dplyr` pipeline that results in two rows and three columns – `year`, `province`, and `median_energy_expense_per_room`.


```{r message=FALSE}
cas %>%
  group_by(year, province) %>%
  summarize(median_energy_expense = median(energy_expense)) %>%
  group_by(year) %>%
  slice_min(median_energy_expense) %>%
  select(year, province, median_energy_expense)

```

## Task 7

7.	Task 7:
A respondent is considered to be “energy vulnerable” if they spend more than 5% of their household income on energy expenses. Recreate the plot below, which shows the proportion of respondents who are energy vulnerable for each combination of age and education. In 2 - 3 sentences, describe what you observe in the plot.
Hints:
o	You will need to use the variable created in task 5.
o	use `geom_raster()`
o	colors are from `"white"` to `"red"` in `scale_fill_gradient()`
o	theme is `bw`
o	figure width is 7, figure height is 6



```
